---
title: "Impact of Opioid Prescription Rate on Suicide Rates and Poverty Levels of Counties in the United States"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning = FALSE,  fig.width =  10)
library(tidyverse)
library(infer)
library(moderndive)
library(cowplot)
```

  

## Introduction

A [recent NPR article](https://www.npr.org/2020/07/17/887590699/doctors-and-dentists-still-flooding-u-s-with-opioid-prescriptions) 
noted that the U.S. is still being flooded with opioid prescriptions by doctors and dentists. 
Upon reading the article, a friend pondered what the provided heat map for prescriptions would look like overlaid 
with suicide rates and poverty levels. This prompted the question "Is a larger rate of legal opioid prescription related to a higher numbers in suicides and poverty?"  

To test the conjecture, we found data from the CDC in relation to 
[suicide rates by county](https://wisqars.cdc.gov:8443/cdcMapFramework/mapModuleInterface.jsp), 
[poverty levels by county](https://data.hrsa.gov/tools/data-explorer?ds=29,34), and 
[opioid prescription rate per county](https://www.cdc.gov/drugoverdose/maps/rxrate-maps.html). 
The suicide rate averages (`Avg_Suicide_Rate`) are calculated per 100,000 people averaged between 2008 and 2014.
Poverty levels (`AvgP_Below_1.00x_FPL`, `AvgP_Below_1.50x_FPL`, `AvgP_Below_2.00x_FPL`) are percentages of families 
in the county below intervals of the Federal Poverty Level (FPL) averaged between 2014 and 2018. We have 
additionally transformed the three intervals into one `Poverty_Index` for the 
county, more heavily weighed towards the percentage of families below the FPL than those above it. The higher the index, 
the more families there are near or in poverty for respective counties. 
The opioid prescription rate averages (`Avg_Prescr_Rate`) are calculated retail opioid prescriptions dispensed 
per 100 persons averaged between 2005 and 2015.  


```{r, label = "import data"}
spo_rate <- read_csv('https://raw.githubusercontent.com/qiqiliang/statistics/master/US_Suicide_Poverty_Opioid_Prescription_Rates') %>%
  select(-X1) %>%

# Adding Poverty_Index
  mutate(Poverty_Index = ((1/3) * (AvgP_Below_1.00x_FPL + AvgP_Below_1.50x_FPL + AvgP_Below_2.00x_FPL))) 

  # The resulting number is a scale from 1 to 100. weighted percentage of people in poverty.
  # This prevents double counting of people and we want those below FPL to weight more heavily on the ranking of a city
  # Calculated by taking AvgP_Below_1.00x_FPL with full weight, AvgP_Below_1.50x_FPL with 2/3 weight, 
  # AvgP_Below_2.00x_FPL with 1/3 weight, and people above AvgP_Below_2.00x_FPL with 0 weight.
  # The higher the poverty index, the more families near or in poverty in respective counties.

```

## Hypotheses  
  
Null Hypothesis: There no correlation between suicide rates, poverty level, and opioid prescription rates.
$$H_0: \quad \beta_S , \beta_P , \beta_O > 0$$
Alternate Hypothesis:  There is correlation between suicide rates, poverty level, and opioid prescription rates. 

  
In the context of our research question, a Type I error represents having some correlation between suicide, 
poverty, and opioid prescription rates but saying otherwise due to an unlikely event in which all values 
have had similar spikes over the years *(this is BS and should totally be looked at)*.
A Type II error would be having no correlation between suicide, poverty, and opioid prescription rates but 
saying otherwise due to...

## Data Exploration  

### Basics
```{r, fig.width = 10, label = "basic histograms"}
o_basic <- ggplot(spo_rate, aes(x = Avg_Prescr_Rate))+
  geom_histogram(binwidth = 10, col = "white") + 
  labs(x = "Opioid Prescription Rate", title = "Opioid Summary Graphic")

s_basic <- ggplot(spo_rate, aes(x = Avg_Suicide_Rate))+
  geom_histogram(binwidth = 1, col = "white")+ 
  labs(x = "Suicide Rate", title = "Suicide Summary Graphic")

pi_basic <- ggplot(spo_rate, aes(x = Poverty_Index))+
  geom_histogram(binwidth = 1, col = "white")+ 
  labs(x = "Poverty Index", title = "Poverty Summary Graphic")
```

```{r, include = FALSE, label = "means and medians"}
spo_rate %>%
  summarise(mn_o_rate = mean(Avg_Prescr_Rate),
            mdn_o_rate = median(Avg_Prescr_Rate),
            mn_s_rate = mean(Avg_Suicide_Rate),
            mdn_s_rate = median(Avg_Suicide_Rate),
            mn_p_rate = mean(Poverty_Index),
            mdn_p_rate = median(Poverty_Index))
```

```{r, label = "basic histograms put together", fig.height = 3}
# arrange the plots to be next to each other
plot_grid(o_basic, s_basic, pi_basic, ncol = 3)
```

The mean opioid prescription rate per 100 persons is 93.63502 while the median is 88.14545, mean suicide rate per 100,000 people is 15.87807 while the median is 14.97576, and the mean poverty index is 19.21391 when looking only at families below two times the FPL whiles the median is 18.82.

*********************************************************************************
  
### Regression models
  it is only appropriate to make predictions based on a regression model if:
    
  1. The data should show a linear trend.
  
  2. The distribution of residuals is approximately Normal.
  
  3. The residuals have constant variability.
  
  
Generate a null distribution for slopes, assuming
  variables independent. Then use `infer` to visualize
  the distribution and compute the p_value. Compare the P-value
  to the one given for slope in the regression table.
  
Generate a bootstrap distribution for slopes. 
    Then use `infer` and the percentile method to create a 95% confidence interval
    on your slope. Compare the confidence interval here to the one given in the
    regression table. Finally, assess whether this claim is reasonable
    given your data set.
    
    
*********************************************************************************


```{r, fig.width=10, label = "regression analysis suicide and opioids"}
so_plot <- ggplot(spo_rate, aes(x = Avg_Prescr_Rate, y = Avg_Suicide_Rate)) +
  geom_point(alpha = 0.3)+
  geom_smooth(method = lm, se = FALSE)+ 
  labs(x = "Opioid Prescription Rate", y = "Suicide Rate", title = "Suicide vs. Opioid Prescription Rates")


m_so <- lm(Avg_Suicide_Rate ~ Avg_Prescr_Rate, data = spo_rate)

m_so_rp <- get_regression_points(m_so) 

so_rp_plot <- ggplot(m_so_rp, aes(x =Avg_Prescr_Rate, y = residual) )+
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE)+ 
   labs(x = "Opioid Prescription Rate", y = "Residual", title = "Scatterplot of Residuals")


so_rp_hplot <- ggplot(m_so_rp, aes(x = residual) )+
  geom_histogram(binwidth = 1, col = "white")+ 
   labs(x = "Opioid Prescription Rate", y = "Residual", title = "Histogram of Residuals")

get_regression_table(m_so)
summary(m_so)

b1_so <- m_so$coefficients[2]

```

```{r, label = "null distribution suicide and opioids"}
so_null <- spo_rate %>% 
  specify(Avg_Suicide_Rate ~ Avg_Prescr_Rate) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "slope") 

so_null_plot <- so_null %>%
  visualize() + 
  shade_p_value(obs_stat = b1_so, direction = "both")+ 
  labs(x = "Slopes", title = "Null Distribution of Slopes with Observed Slope")

# so_null %>%
#   get_p_value(obs_stat = b1_so, direction = "both")
```

```{r, label = "suicide and opioid plots"}
# arrange the plots to be next to each other
pdf(width = 10, "~/Documents/Stats Project w: qiqi/statistics/plots and figures/SR_OPR.pdf")
plot_grid(so_plot, so_rp_plot, so_rp_hplot, so_null_plot, ncol = 2)
dev.off()
```

```{r, label = "boostrap ci suicide and opioids", include = FALSE}
so_boot <- spo_rate %>% 
  specify(Avg_Suicide_Rate ~ Avg_Prescr_Rate) %>%
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "slope") 

so_boot %>%
  get_ci(level = 0.95, type = "percentile")
```

Opioid prescription rate may be a statistically significant predictor of suicide rate as the p value 
calculated is less than the significance level $\alpha = 0.05$. The linear model, however, does not 
seems to be too practical, as the adjusted $R^2 = 0.06084$.
The calculated slope is $\beta_1 = 0.031532$ with a confidence interval of $(0.026, 0.037)$.

***This next part tells us whether it's appropriate to make assumptions off the regression model based on the guidelines above.***   

***The real question is, do we need to make assumptions off it and what does that mean? We're just looking for any relationship.***

The histogram of residuals is slightly right skewed, but still uni-modal and close to 0. This shows constant variability, a linear trend, and approximate normalicy. 

```{r, fig.width=10, label = "regression analysis 1xFPL and opioids"}
sp1_plot <- ggplot(spo_rate, aes(x = Avg_Prescr_Rate, y = AvgP_Below_1.00x_FPL)) +
  geom_point(alpha = 0.3)+
  geom_smooth(method = lm, se = FALSE)+ 
  labs(x = "Opioid Prescription Rate", y = "% Families Below 1X FPL", title = "% Families Below 1X FPL vs. Opioid Prescription Rate")


m_sp1 <- lm(AvgP_Below_1.00x_FPL ~ Avg_Prescr_Rate, data = spo_rate)

m_sp1_rp <- get_regression_points(m_sp1) 

sp1_rp_plot <- ggplot(m_sp1_rp, aes(x =Avg_Prescr_Rate, y = residual) )+
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE)+ 
  labs(x = "Opioid Prescription Rate", y = "Residual", title = "Scatterplot of Residuals")


sp1_rp_hplot <- ggplot(m_sp1_rp, aes(x = residual) )+
  geom_histogram(binwidth = 1, col = "white")+ 
  labs(x = "Opioid Prescription Rate", y = "Residual", title = "Histogram of Residuals")

# get_regression_table(m_sp1)
# summary(m_sp1)

b1_sp1 <- m_sp1$coefficients[2]

```

```{r, label = "null distribution 1xFPL and opioids"}
sp1_null <- spo_rate %>% 
  specify(AvgP_Below_1.00x_FPL ~ Avg_Prescr_Rate) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "slope") 

sp1_null_plot <- sp1_null %>%
  visualize() + 
  shade_p_value(obs_stat = b1_sp1, direction = "both")+ 
  labs(x = "Slopes", title = "Null Distribution of Slopes with Observed Slope")

# sp1_null %>%
#   get_p_value(obs_stat = b1_sp1, direction = "both")
```

```{r, label = "1xFPL and opioid plots"}
# arrange the plots to be next to each other
pdf(width = 10, "~/Documents/Stats Project w: qiqi/statistics/plots and figures/P1_OPR.pdf")
plot_grid(sp1_plot, sp1_rp_plot, sp1_rp_hplot, sp1_null_plot, ncol = 2)
dev.off()
```

```{r, label = "boostrap ci 1xFPL and opioids", include = FALSE}
sp1_boot <- spo_rate %>% 
  specify(AvgP_Below_1.00x_FPL ~ Avg_Prescr_Rate) %>%
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "slope")

sp1_boot %>%
  get_ci(level = 0.95, type = "percentile")
```
Opioid prescription rate may be a statistically significant predictor of the percentage of 
families below 1x the FPL as the p value calculated is less than the significance level 
$\alpha = 0.05$. The linear model, however, does not seems to be too practical, as the adjusted $R^2 = 0.1608$.
The calculated slope is $\beta_1 = 3.5944$ with a confidence interval of $(3.21, 3.98)$. 

```{r, fig.width=10, label = "regression analysis 1.50XFPL and opioids"}
sp1_5_plot <- ggplot(spo_rate, aes(x = Avg_Prescr_Rate, y = AvgP_Below_1.50x_FPL)) +
  geom_point(alpha = 0.3)+
  geom_smooth(method = lm, se = FALSE)+ 
  labs(x = "Opioid Prescription Rate", y = "% Families Below 1.50X FPL", title = "% Families Below 1.50X FPL vs. Opioid Prescription Rate")


m_sp1_5 <- lm(AvgP_Below_1.50x_FPL ~ Avg_Prescr_Rate, data = spo_rate)

m_sp1_5_rp <- get_regression_points(m_sp1_5) 

sp1_5_rp_plot <- ggplot(m_sp1_5_rp, aes(x =Avg_Prescr_Rate, y = residual) )+
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE)+ 
  labs(x = "Opioid Prescription Rate", y = "Residual", title = "Scatterplot of Residuals")


sp1_5_rp_hplot <- ggplot(m_sp1_5_rp, aes(x = residual) )+
  geom_histogram(binwidth = 1, col = "white")+ 
  labs(x = "Opioid Prescription Rate", y = "Residual", title = "Histogram of Residuals")

# get_regression_table(m_sp1_5)
# summary(m_sp1_5)

b1_sp1_5 <- m_sp1_5$coefficients[2]

```

```{r, label = "null distribution 1.50XFPL and opioids"}
sp1_5_null <- spo_rate %>% 
  specify(AvgP_Below_1.50x_FPL ~ Avg_Prescr_Rate) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "slope") 

sp1_5_null_plot <- sp1_5_null %>%
  visualize() + 
  shade_p_value(obs_stat = b1_sp1_5, direction = "both")+ 
  labs(x = "Slopes", title = "Null Distribution of Slopes with Observed Slope")

# sp1_5_null %>%
#   get_p_value(obs_stat = b1_sp1_5, direction = "both")
```

```{r, label = "1.50XFPL and opioid plots"}
# arrange the plots to be next to each other
pdf(width = 10, "~/Documents/Stats Project w: qiqi/statistics/plots and figures/P1_5_OPR.pdf")
plot_grid(sp1_5_plot, sp1_5_rp_plot, sp1_5_rp_hplot, sp1_5_null_plot, ncol = 2)
dev.off()
```

```{r, label = "boostrap ci 1.5xFPL and opioids", include= FALSE}
sp1_5_boot <- spo_rate %>% 
  specify(AvgP_Below_1.50x_FPL ~ Avg_Prescr_Rate) %>%
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "slope")

sp1_5_boot %>%
  get_ci(level = 0.95, type = "percentile")
```

Opioid prescription rate may be a statistically significant predictor of the percentage of families 
below 1.5x the FPL as the p value calculated is less than the significance level $\alpha = 0.05$. 
The linear model, however, does not seems to be too practical, as the adjusted $R^2 = 0.159$.
The calculated slope is $\beta_1 = 2.380$ with a confidence interval of $(2.12, 2.64)$. 

```{r, fig.width=10, label = "regression analysis 2XFPL and opioids"}
sp2_plot <- ggplot(spo_rate, aes(x = Avg_Prescr_Rate, y = AvgP_Below_2.00x_FPL)) +
  geom_point(alpha = 0.3)+
  geom_smooth(method = lm, se = FALSE)+ 
  labs(x = "Opioid Prescription Rate", y = "% Families Below 2X FPL", title = "% Families Below 2X FPL vs. Opioid Prescription Rate")


m_sp2 <- lm(AvgP_Below_2.00x_FPL ~ Avg_Prescr_Rate, data = spo_rate)

m_sp2_rp <- get_regression_points(m_sp2) 

sp2_rp_plot <- ggplot(m_sp2_rp, aes(x =Avg_Prescr_Rate, y = residual) )+
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE)+ 
  labs(x = "Opioid Prescription Rate", y = "Residual", title = "Scatterplot of Residuals")


sp2_rp_hplot <- ggplot(m_sp2_rp, aes(x = residual) )+
  geom_histogram(binwidth = 1, col = "white")+ 
  labs(x = "Opioid Prescription Rate", y = "Residual", title = "Histogram of Residuals")

# get_regression_table(m_sp2)
# summary(m_sp2)

b1_sp2 <- m_sp2$coefficients[2]

```

```{r, label = "null distribution 2XFPL and opioids"}
sp2_null <- spo_rate %>% 
  specify(AvgP_Below_2.00x_FPL ~ Avg_Prescr_Rate) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "slope") 

sp2_null_plot <- sp2_null %>%
  visualize() + 
  shade_p_value(obs_stat = b1_sp2, direction = "both")+ 
  labs(x = "Slopes", title = "Null Distribution of Slopes with Observed Slope")

# sp2_null %>%
#   get_p_value(obs_stat = b1_sp2, direction = "both")
```

```{r, label = "2XFPL and opioid plots"}
# arrange the plots to be next to each other
pdf(width = 10, "~/Documents/Stats Project w: qiqi/statistics/plots and figures/P2_OPR.pdf")
plot_grid(sp2_plot, sp2_rp_plot, sp2_rp_hplot, sp2_null_plot, ncol = 2)
dev.off()
```

```{r, label = "boostrap ci 2xFPL and opioids", include = FALSE}
sp2_boot <- spo_rate %>% 
  specify(AvgP_Below_2.00x_FPL ~ Avg_Prescr_Rate) %>%
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "slope")

sp2_boot %>%
  get_ci(level = 0.95, type = "percentile")
```

Opioid prescription rate may be a statistically significant predictor of the percentage of families 
below 2.0x the FPL as the p value calculated is less than the significance level $\alpha = 0.05$. 
The linear model, however, does not seems to be too practical, as the adjusted $R^2 = 0.1632$.
The calculated slope is $\beta_1 = 0.086358$ with a confidence interval of $(0.077, 0.095)$. 

```{r, fig.width=10, label = "regression analysis poverty index and opioids"}
spi_plot <- ggplot(spo_rate, aes(x = Avg_Prescr_Rate, y = Poverty_Index)) +
  geom_point(alpha = 0.3)+
  geom_smooth(method = lm, se = FALSE)+ 
  labs(x = "Opioid Prescription Rate", y = "Poverty Index", title = "Poverty Index vs. Opioid Prescription Rate")


m_spi <- lm(Poverty_Index ~ Avg_Prescr_Rate, data = spo_rate)

m_spi_rp <- get_regression_points(m_spi) 

spi_rp_plot <- ggplot(m_spi_rp, aes(x =Avg_Prescr_Rate, y = residual) )+
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE)+ 
  labs(x = "Opioid Prescription Rate", y = "Residual", title = "Scatterplot of Residuals")


spi_rp_hplot <- ggplot(m_spi_rp, aes(x = residual) )+
  geom_histogram(binwidth = 1, col = "white")+ 
  labs(x = "Opioid Prescription Rate", y = "Residual", title = "Histogram of Residuals")

get_regression_table(m_spi)
summary(m_spi)

b1_spi <- m_spi$coefficients[2]

```

```{r, label = "null distribution poverty index and opioids"}
spi_null <- spo_rate %>% 
  specify(Poverty_Index ~ Avg_Prescr_Rate) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "slope") 

spi_null_plot <- spi_null %>%
  visualize() + 
  shade_p_value(obs_stat = b1_spi, direction = "both")+ 
  labs(x = "Slopes", title = "Null Distribution of Slopes with Observed Slope")

# spi_null %>%
#   get_p_value(obs_stat = b1_spi, direction = "both")
```

```{r, label = "poverty index and opioid plots"}
# arrange the plots to be next to each other
pdf("~/Documents/Stats Project w: qiqi/statistics/plots and figures/PI_OPR.pdf", width = 10)
plot_grid(spi_plot, spi_rp_plot, spi_rp_hplot, spi_null_plot, ncol = 2)
dev.off()
```

```{r, label = "boostrap ci poverty index and opioids", include= FALSE}
spi_boot <- spo_rate %>% 
  specify(Poverty_Index ~ Avg_Prescr_Rate) %>%
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "slope")

spi_boot %>%
  get_ci(level = 0.95, type = "percentile")
```



Opioid prescription rate may be a statistically significant predictor of our calculated poverty index as the p value calculated 
is less than the significance level $\alpha = 0.05$. 
The linear model, however, does not seems to be too practical, as the adjusted $R^2 = 0.1652$.
The calculated slope is $\beta_1 = 0.066072$ with a confidence interval of $(0.059, 0.073)$. 





If the null hypothesis were true, then the observed slope does not seem plausible because it is an extreme value rather than a more expected value near the center of the distribution. The P-value, assuming a significance level of $\alpha = 0.05$, is 0 when rounded to the thousandth, which means we can reject the null-hypothesis and test for aspects of the alternative. The P-value here and the one generated by the linear regression above is the same when rounded to the thousandth.


## Quick Summary  

The mean opioid prescription rate per 100 persons is 93.63502 while the median is 88.14545, mean suicide rate per 100,000 people is 15.87807 while the median is 14.97576, and the mean poverty index is 19.21391 when looking only at families below two times the FPL whiles the median is 18.82.



Variable | Mean | Median | Units
---------|------|------- | ---------
Opioid Prescription | 93.63502 | 93.63502 |  Prescriptions per 100 persons
Suicide | 15.87807 | 14.97576 | Suicides per 100,000 people
Poverty Index | 19.21391 | 18.82 | N/A 
  
Model | Null Distribution P-Value | LM P-Value | Adjusted $R^2$ | Slope | LM Slope CI | Bootstrap CI  
--|--|--|--|--|--|--
OPR x SR | 0 | < 2.2e-16 | 0.06084 | 0.031532 | (0.026, 0.037) | (0.02573316, 0.03740757)  
OPR x P1.0 | 0 | < 2.2e-16 | 0.1608 | 0.044857 | (0.040, 0.050) | (0.03914246, 0.05022237)  
OPR x P1.5 | 0 | < 2.2e-16 | 0.159 | 0.067001 | (0.060. 0.074) | (0.05913409, 0.07493694) 
OPR x P2.0 | 0 | < 2.2e-16 | 0.1632 | 0.086358 | (0.077, 0.095) | (0.07714104, 0.09526279)  
OPR x PI | 0 | < 2.2e-16 | 0.1652 | 0.066072 | (0.059, 0.073) | (0.05869566, 0.07321176)



*************************


## Let's Put Stuff Together  
  
Now consider how the association between suicide rate and opioid prescription rate would change if poverty brackets were taken into account. Create a plot to illustrate the
    relationship between all three variables (recall the use of the
    `color` argument in `aes()`). Just by looking at the plot and
    without fitting any models, does it appear that the relationship
    between the suicide rate and opioid prescription rate is the same for higher poverty counties as it is
    for lower poverty counties?

```{r}
mid <- mean(spo_rate$Poverty_Index)

# Color by poverty index
spo_plot <-ggplot(spo_rate, aes(x = Avg_Prescr_Rate, y = Avg_Suicide_Rate, color=Poverty_Index)) + 
  geom_point(position = "jitter") + 
  # geom_smooth(method = "lm", se = FALSE)+ 
  scale_color_gradient2(midpoint=mid, low="green", mid="white", high="red", space ="Lab", name = "Poverty Index\n (0 to 100)") +
  labs(title = "Visualization of Opioids, Suicide, and Poverty Relationship", y = "Suicides (per 100,000 people)", x = "Opioid Prescriptions (per 100 persons)")

pdf(width = 10, "~/Documents/Stats Project w: qiqi/statistics/plots and figures/SPO.pdf")
spo_plot
dev.off()
```

*********************************************************************************






```{r, fig.width=10, label = "regression analysis poverty index and SUICIDE"}
pis_plot <- ggplot(spo_rate, aes(x = Avg_Suicide_Rate, y = Poverty_Index)) +
  geom_point(alpha = 0.3)+
  geom_smooth(method = lm, se = FALSE)+ 
  labs(x = "Suicide Rate", y = "Poverty Index", title = "Poverty Index vs. Suicide Rate")


m_pis <- lm(Poverty_Index ~ Avg_Suicide_Rate, data = spo_rate)

m_pis_rp <- get_regression_points(m_pis) 

pis_rp_plot <- ggplot(m_pis_rp, aes(x =Avg_Suicide_Rate, y = residual) )+
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE)+ 
  labs(x = "Suicide Rate", y = "Residual", title = "Scatterplot of Residuals")


pis_rp_hplot <- ggplot(m_pis_rp, aes(x = residual) )+
  geom_histogram(binwidth = 1, col = "white")+ 
  labs(x = "Suicide Rate", y = "Residual", title = "Histogram of Residuals")

# get_regression_table(m_pis)
# summary(m_pis)

b1_pis <- m_pis$coefficients[2]

```

```{r, label = "null distribution poverty index and SUICIDE"}
pis_null <- spo_rate %>% 
  specify(Poverty_Index ~ Avg_Suicide_Rate) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "slope") 

pis_null_plot <- pis_null %>%
  visualize() + 
  shade_p_value(obs_stat = b1_pis, direction = "both")+ 
  labs(x = "Slopes", title = "Null Distribution of Slopes with Observed Slope")

pis_null %>%
  get_p_value(obs_stat = b1_pis, direction = "both")
```

```{r, label = "poverty index and SUICIDE plots"}
# arrange the plots to be next to each other
pdf(width = 10, "~/Documents/Stats Project w: qiqi/statistics/plots and figures/SR_PI.pdf")
plot_grid(pis_plot, pis_rp_plot, pis_rp_hplot, pis_null_plot, ncol = 2)
dev.off()
```

```{r, label = "boostrap ci poverty index and SUICIDE"}
pis_boot <- spo_rate %>% 
  specify(Poverty_Index ~ Avg_Suicide_Rate) %>%
  generate(reps = 1000, type = "bootstrap") %>% 
  calculate(stat = "slope")

pis_boot %>%
  get_ci(level = 0.95, type = "percentile")
```

<!-- Backward-selection using p-value is a model selection process in which -->
<!-- we start with the full model and consecutively remove the variable  -->
<!-- from the model with highest p-value and refit the model, until all  -->
<!-- variables have p-value at most 0.05 (or another discipline specific -->
<!-- significance level), -->

<!-- Using backward-selection and p-value as the selection criterion, -->
<!-- determine the best model. Write out the -->
<!-- linear model for predicting score based on the final model you -->
<!-- settle on. -->


```{r, include = FALSE, label = "multiple regression analysis poverty levels, suicide, and opioids"}
m_full <- lm(Avg_Prescr_Rate ~ Poverty_Index + AvgP_Below_2.00x_FPL + 
               AvgP_Below_1.50x_FPL + AvgP_Below_1.00x_FPL + 
               Avg_Suicide_Rate, data = spo_rate)

get_regression_table(m_full)
get_regression_summaries(m_full)
summary(m_full)
```

<!-- Drop the variable with the highest p-value and re-fit the model. -->

```{r, include = FALSE}
m_no_1_5 <- lm(Avg_Prescr_Rate ~ AvgP_Below_2.00x_FPL + AvgP_Below_1.00x_FPL + Avg_Suicide_Rate, data = spo_rate)

get_regression_table(m_full)
get_regression_summaries(m_full)
```

<!-- Drop the variable with the highest p-value and re-fit the model. -->

```{r, include = FALSE}
m_only_1 <- lm(Avg_Prescr_Rate ~ AvgP_Below_1.00x_FPL + Avg_Suicide_Rate, data = spo_rate)

get_regression_table(m_full)
get_regression_summaries(m_full)
```

<!-- $$\begin{aligned} \widehat{Avg\_Prescr\_Rate} &= 30.648 + 3.410 \times AvgP\_Below\_1.00x\_FPL  -->
<!-- \\&+ 1.653 \times Avg\_Suicide\_Rate \end{aligned}$$ -->


```{r, include = FALSE, label = "multiple regression analysis poverty index, suicide, and opioids"}
m_spi <- lm(Avg_Prescr_Rate ~ Poverty_Index + 
               Avg_Suicide_Rate, data = spo_rate)

get_regression_table(m_spi)
get_regression_summaries(m_spi)
```

<!-- $$\begin{aligned} \widehat{Avg\_Prescr\_Rate} &= 25.853	 + 2.320 \times Poverty\_Index  -->
<!-- \\&+ 1.461 \times Avg\_Suicide\_Rate \end{aligned}$$ -->


<!-- The interpretation of the coefficients in multiple regression is -->
<!-- slightly different from that of simple regression. The estimate for -->
<!-- `AvgP_Below_1.00x_FPL` reflects how many more prescriptions a county is  -->
<!-- expected to have received if they have percentage of families under two  -->
<!-- times the FPL that is one point higher *while -->
<!-- holding all other variables constant*. In this case, that translates -->
<!-- into considering only counties of the same rank with `AvgP_Below_1.00x_FPL` scores -->
<!-- that are one percent apart. -->

*********************************************************************************

<!-- More questions to ask: -->
<!--  - Demographics???? -->
<!--     - Ethnicity, Income -->
<!--     - Targeting marginalized communities? -->
<!--     - access to healthcare -->
